#**Course Project 1**  
*This markdown document is for Course Project 1 of Reproducible Research Course.  

The following tasks are carried out for this project:  

*Code for reading the dataset and/or processing the data* 

#__Loading and Preprocessing of the Data__  
```{r echo = TRUE}
data <- read.csv("activity.csv", header = TRUE, sep =",")

```

- to check how many NAs and which column contains NAs
```{r echo = TRUE}
dim(data)
table(is.na(data))
table(is.na(data$steps))
table(is.na(data$date))
table(is.na(data$interval))
```
Only the variable, steps contain 2304 NAs out of 17568 observations.  

#__What is mean total number of steps taken per day?__   
```{r echo = TRUE}
steps_perday <- aggregate(steps ~ date, data, FUN = sum, na.rm = TRUE)

png("plot1.png", width = 480, height = 480, unit = "px") #Set up the png plot
hist(steps_perday$steps, main = "Total number of steps taken each day", 
     xlab = "Number of steps", ylab = "Frequency")

dev.off()

#Mean and median of the total numer of steps taken per day
summary(steps_perday$steps, na.rm = TRUE)

```
- The __mean__ of the total number of steps taken per day = 10766.  

- The __median__ of the total number of steps taken per day = 10765.  
- Can also use the mean and median functions


#__What is the average daily activity pattern?__  

```{r echo = TRUE}
#loading the data 
#data <- read.csv("activity.csv", header = TRUE, sep =",")
#dim(data)
#head(data)

#Plot the activity graph using line graph (type = "l")
avg_pattern <- aggregate(steps~interval, data, mean, na.rm = TRUE)

png("plo2.png", width = 480, height = 480, unit = "px") #Set up the png plot
plot(avg_pattern, type = "l", col = "red", 
     main = "The daily activity pattern", 
     xlab = "5-minute interval", ylab = "Average number of steps")

dev.off()

```

#*Which 5-minute interval, on average across all the dataset, contains the maximum number of steps?*  
```{r echo = TRUE}
max_interval <- avg_pattern$interval[which.max(avg_pattern$steps)]
```
The 835th interval contains the maximum number of steps   


#__Imputing missing values__  


#head(data)
```{r echo = TRUE} 
#1. To calculate and report the total number of missing values 
total_NA <- sum(is.na(data))
total_NA #Total missing data
sum(is.na(data$steps))#missing data in 'steps' column = 2304
sum(is.na(data$date)) #missing data in "date' column = 0
table(is.na(data$interval)) #missing in 'interval' = 0

#The missing values will be imputed based on mean and since only one column
#contains the missing data, it is pretty straightforward 
head(data) #The top 6 columns of 'steps' contain NAs

#Impute NA in steps and generate a new dataframe 'clean_data
clean_data <- data
clean_data$steps[is.na(clean_data$steps)] <- mean(data$steps, na.rm = TRUE)
head(data)  #For comparison
tail(data)

head(clean_data) #The NAs are replaced by the mean = 37.3826
tail(clean_data)

#to plot a histogram of the total number of steps taken each day
no_of_steps <- aggregate(steps~date, clean_data, FUN = sum)
png("Plot3.png")
par(mfrow = c(2,1))
hist(no_of_steps$steps, col = "blue", ylim = c(0,35),
     main = "Total number of steps taken each day (after imupting NAs)",
     xlab = "Total number of steps", ylab = "Frequency")

#To plot original data on the righ side
org_data <- aggregate(steps ~ date, data, FUN = sum, na.rm = TRUE)
hist(org_data$steps, col = "green", ylim = c(0,35),
     main = "Total number of steps taken each day (Original data)",
     xlab = "Total number of steps", ylab = "Frequency")
dev.off()
# From the figures, there is a difference after imputing the NAs. 

#To check the actual difference, we can do the following calculation
org_mean <- mean(org_data$steps, na.rm = TRUE)
org_median <- median(org_data$steps, na.rm = TRUE)
org_totalSteps <- sum(org_data$steps)

new_mean <- mean(no_of_steps$steps, na.rm = TRUE)
new_median <- median(no_of_steps$steps, na.rm = TRUE)
new_totalSteps <- sum(no_of_steps$steps)

#Differences 
diff_mean <- org_mean - new_mean
diff_median <- org_median - new_median
diff_totalSteps <- org_totalSteps - new_totalSteps
cat("The difference in mean = ", diff_mean, "\nThe difference in median is = ", diff_median, 
    "\n The total difference in number of steps = ", diff_totalSteps)


```

__Are there differences in activity patterns between weekdays and weekends?__  

*Set a function to segregate the weekends and weekdays and then add as variable
```{r echo = TRUE}
#Using the weekdays() create weekend and weekday
daysofweek <- function(date){
  day <- weekdays(date)
  if(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("Weekday")
           else if (day %in% c("Saturday", "Sunday"))
             return ("Weekend")
  else
    stop("Invalid Date")
}

clean_data$steps[is.na(data$steps)] <- mean(data$steps, na.rm = TRUE)
final_data <- aggregate(steps ~ interval + date, clean_data, FUN = sum)
final_data$date <- as.Date(final_data$date)
final_data$day <- sapply(final_data$date, FUN = daysofweek)
head(final_data)
tail(final_data) #data looks correct

mean_stepsperDay <- aggregate(steps ~ interval+day, final_data, mean)

png("Plot4.png")

library(ggplot2)
g <- ggplot(mean_stepsperDay, aes(x = interval, y = steps))
g + geom_line(size = 1) + facet_grid(day ~.) +
ggtitle("Average daily activity time series") +          
xlab("Interval") + 
ylab("Number of steps")
dev.off()

```
